### 2025-01-28 paper/icml25/first/sects/background.tex
minors

\subsection{Markov Decision Processes}
To provide context for our work, we first introduce the reinforcement learning (RL) framework. Each task is modeled as a partially observable, decentralized, finite-horizon, Markov Decision Process (MDP)~\cite{Shapley53}, $\mdp$: $(\agents,\states,\allo,\alla,\allp, \allr, \transition, \horizon, \discount, \initdist)$. Where $\agents=[\na]$ and $\states$ denote the set of agents and the state space, respectively. $\allo={\{\observations{i}\}}_{i\in\agents}$ and $\alla={\{\actions{i}\}}_{i\in\agents}$ are the observation space and action space for each agent. ${\{\stateobs{i}:\states\rightarrow\observations{i}\}}_{i\in\agents}$ are the projections from the state to the observation space and ${\{\rewards{i}:\states\times\actions{i}\times\states\rightarrow\mathbb{R}\}}_{i\in\agents}$ are the reward functions for each agent. $\transition:\states\times\alla$ denotes the transition function. $\discount$ and $\horizon$ denote the discount factor and the time horizon respectively. Finally, $\initdist$ is the initial state distribution.

@@

A policy for agent $i$ is a function $\policy{i}:\observations{i}\rightarrow\actions{i}$ that maps its observation space to its action space. We will denote the collective policy, determined by the collection of agent-specific policies simply with $\pi$. We can use the policy to sample trajectories from the MDP by iteratively applying the transition, reward, projection and policy functions. In this case, for trajectory $\tau$, we have that the $i$-th action for the $j$-th agent is $a_{ij}^\tau = \policy{j}(\stateobs{j}(s_i^\tau))$ and the next state is $s_{i+1}^\tau = \transition(s_i^\tau,a_{i1},\dots,a_{i\na})$. The $i$-th reward for the $j$-th agent is $r_{ij}^\tau = \rewards{j}(s_i^\tau,a_{ij}^\tau,s_{i+1}^\tau)$. Finally, $s_0$ is sampled from the initial state distribution $\initdist$. For brevity, we will simply denote with $\tau\sim\pi$ a trajectory sampled following the policies.

@@

In its trivial multi-agent extension, SHAC trains policies ${\pi_1}_\theta,\dots,{\pi_\na}_\theta$ by minimizing the loss function $\mathcal{L}_\theta^{\pi}:\trajectories^\bs\rightarrow\mathbb{R}$ for a batch of $\bs$ trajectories $\batch = \{\tau_1, \ldots, \tau_\bs\}$ sampled from $\mdp$. The loss $\mathcal{L}_\theta^{\pi}(\batch)$ is defined as:  
\begin{equation*}\label{eq:shac}
    \frac{1}{\bs\na\mh}\sum_{\tau\in\batch} \sum_{t=t_0}^{t_0+\mh-1} \sum_{i\in\agents} \discount^{t-t_0} R(s_{t}^\tau, a_{ti}^\tau) + \discount^\mh \valuef_\theta(s^\tau_{t_0+\mh})
\end{equation*}
where $\mh \ll \horizon$ is the rollout size, $t_0$ is the initial time step of the rollout, $\valuef_\theta: \states \rightarrow \mathbb{R}$ is the value function estimating the expected return. It is apparent that SHAC assumes that the policy, reward, value, projection, and transition functions are all differentiable, i.e., $\{\policy{i}\}_{i\in\agents},\{\rewards{i}\}_{i\in\agents},\valuef_\theta,\transition \in \mathcal{C}^1$. While this is a common requirement for policy and value functions, differentiable transition and reward functions are rarely available.

>>>>

\subsection{Markov Decision Processes}
To provide context for our work, we first introduce the reinforcement learning (RL) framework. Each task is modeled as a partially observable, decentralized, finite-horizon, Markov Decision Process (MDP)~\cite{Shapley53}, $\mdp$: $(\agents,\states,\allo,\alla,\allp, \allr, \transition, \horizon, \discount, \initdist)$. Where $\agents=[\na]$ and $\states$ denote the set of agents and the state space, respectively. $\allo={\{\observations{i}\}}_{i\in\agents}$ and $\alla={\{\actions{i}\}}_{i\in\agents}$ are the observation space and action space for each agent. ${\{\stateobs{i}:\states\rightarrow\observations{i}\}}_{i\in\agents}$ are the projections from the state to the observation space and ${\{\rewards{i}:\states\times\actions{i}\times\states\rightarrow\mathbb{R}\}}_{i\in\agents}$ are the reward functions for each agent. $\transition:\states\times\alla\rightarrow\states$ denotes the transition function. $\discount$ and $\horizon$ denote the discount factor and the time horizon respectively. Finally, $\initdist$ is the initial state distribution.

@@

A policy for agent $i$ is a function $\policy{i}:\observations{i}\rightarrow\actions{i}$ that maps its observation space to its action space. We will denote the collective policy, determined by the collection of agent-specific policies simply with $\pi=\{\pi_{i}\}_{i\in\agents}$. We can use the policy to sample trajectories from the MDP by iteratively applying the transition, reward, projection and policy functions. In this case, for trajectory $\tau$, we have that the $i$-th action for the $j$-th agent is $a_{ij}^\tau = \policy{j}(\stateobs{j}(s_i^\tau))$ and the next state is $s_{i+1}^\tau = \transition(s_i^\tau,a_{i1},\dots,a_{i\na})$. The $i$-th reward for the $j$-th agent is $r_{ij}^\tau = \rewards{j}(s_i^\tau,a_{ij}^\tau,s_{i+1}^\tau)$. Finally, $s_0$ is sampled from the initial state distribution $\initdist$. For brevity, we will simply denote with $\tau\sim\pi$ a trajectory sampled following the policies $\pi$.

@@

In its trivial multi-agent extension, SHAC trains policies ${\pi_1}_\theta,\dots,{\pi_\na}_\theta$ by minimizing the loss function $\mathcal{L}_\theta^{\pi}:\trajectories^\bs\rightarrow\mathbb{R}$ for a batch of $\bs$ trajectories $\batch = \{\tau_1, \ldots, \tau_\bs\}$ sampled usin $\pi$. The loss $\mathcal{L}_\theta^{\pi}(\batch)$ is defined as:  
\begin{equation*}\label{eq:shac}
    \frac{1}{\bs\na\mh}\sum_{\tau\in\batch} \sum_{t=t_0}^{t_0+\mh-1} \sum_{i\in\agents} \discount^{t-t_0} R(s_{t}^\tau, a_{ti}^\tau) + \discount^\mh \valuef_\theta(s^\tau_{t_0+\mh})
\end{equation*}
where $\mh \ll \horizon$ is the rollout size, $t_0$ is the initial time step of the rollout, $\valuef_\theta: \states \rightarrow \mathbb{R}$ is the value function estimating the expected return. It is apparent that SHAC assumes that the policy, reward, value, projection, and transition functions are all differentiable, i.e., $\{\policy{i}\}_{i\in\agents},\{\rewards{i}\}_{i\in\agents},\valuef_\theta,\transition \in \mathcal{C}^1$. While this is a common requirement for policy and value functions, differentiable transition and reward functions are rarely available.

### 2025-01-28 paper/icml25/first/sects/discussion.tex
minors

This advantage does not hold in the Dispersion scenario, where complexity increases substantially with additional agents. Nevertheless, we observe the emergence of cooperation across all scenarios, suggesting potential scalability to larger agent populations. The main bottleneck lies in the transformer architecture for the action world model, where the sequence length scales with the number of agents (sequence length = num.\ agents times num.\ train steps). With significantly large populations (~50 agents), the attention mechanism becomes computationally expensive, leading to prohibitively large memory requirements. In these scenarios, a linear attention mechanism (such as~\cite{Beltagy20}) may be employed.

>>>>

This advantage does not hold in the Dispersion scenario, where complexity increases substantially with additional agents. Nevertheless, we observe the emergence of cooperation across all scenarios, suggesting potential scalability to larger agent populations. The main bottleneck lies in the transformer architecture for the action world model, where the sequence length scales with the number of agents (sequence length = num.\ agents times num.\ train steps). With significantly large populations ($\sim$50 agents), the attention mechanism becomes computationally expensive, leading to prohibitively large memory requirements. In these scenarios, a linear attention mechanism (such as~\cite{Beltagy20}) may be employed.

### 2025-01-28 paper/icml25/first/sects/shacpp.tex
few corrections...
minors

\begin{figure*}[t]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/shac.pdf}
        \caption{SHAC}
        \label{fig:shac}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/shacpp.pdf}
        \caption{\fname{}}
        \label{fig:shacpp}
    \end{subfigure}
    \caption{Comparison between SHAC and \fname{}. Nodes with labels (such as $r_i$ or $r'_i$ represents values). Nodes without labels (such as $\pi$ or $R$) represents functions. Dashed lines are used to represent non differentiability. $a_i$ and $r_i$ represent the collection of actions and rewards for all agents}\label{fig:shac-shacpp}
\end{figure*}

@@

Similarly, we aim to train a neural network to approximate the transition function. However, to avoid vanishing or exploding gradients, we employ a non-recursive approach similar to Action World Model~\cite{Ma24}. This corresponds to train a neural network that given a starting observation from $\mathcal{S}$ and a sequence of $\mh$ actions from $\alla^\mh$, predicts the next $\mh$ observations from $\states^\mh$. We aim to minimize:
\begin{align*}
    \underset{\tau\sim\pi}{E}\left[\hspace{-1.2cm}\sum_{\scriptscriptstyle \hspace{1cm}\substack{s_t,\dots,s_{t+h+1},\\\hspace{.3cm}a_{t*},\dots,a_{t+h*}\in\tau}}\hspace{-1.2cm} \left\vert\left\vert \transition_{\theta}(s_t, a_{t*},\dots,a_{\scriptscriptstyle i+h*}) - [s_t,\dots,s_{\scriptscriptstyle t+h+1}] \right\vert\right\vert^2\right] 
\end{align*}
Where $a_{t*}=\{a_{ti}\}_{i\in\agents}$. In practice, this is achieved by caching trajectories and then training the neural network to minimize the MSE loss. That is, we minimize the loss $\mathcal{L}_\theta^{F}:\mathcal{T}^N\rightarrow\mathbb{R}$:
$$ \frac{1}{\bs\mh}\sum_{\tau \in B}\left|\left| \transition_{\theta}(s_0^\tau, a_{1*}^\tau, \dots, a_{\mh*}) - \left[ s^\tau_1, \dots, s^\tau_\mh \right] \right|\right|^2 $$
Where $B$ represents a batch of trajectories of length $\mh$ sampled uniformly from the cache.

@@

$$\mathcal{L}_\theta^\pi(\batch) + \alpha \frac{1}{\bs\horizon}\sum_{\tau\in\batch}\sum_{a\in\tau} d(a^\tau,(a,b))$$
$$d(x,(l,r)) = \begin{cases}(l-x)^2 & \text{if } x < l \\ (x-r)^2 & \text{if } x > r \\ 0 & \text{otherwise} \end{cases}$$

>>>>

\begin{figure*}[t]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/shac.pdf}
        \caption{SHAC}%
        \label{fig:shac}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/shacpp.pdf}
        \caption{\fname{}}%
        \label{fig:shacpp}
    \end{subfigure}
    \caption{Comparison between SHAC and \fname{}. Nodes with labels (such as $r_i$ or $r'_i$ represents values). Nodes without labels (such as $\pi$ or $R$) represents functions. Dashed lines are used to represent non differentiability. $a_i$ and $r_i$ represent the collection of actions and rewards for all agents}\label{fig:shac-shacpp}
\end{figure*}

@@

Similarly, we aim to train a neural network to approximate the transition function. However, to avoid vanishing or exploding gradients, we employ a non-recursive approach similar to Action World Model~\cite{Ma24}. This corresponds to train a neural network that given a starting observation from $\mathcal{S}$ and a sequence of $\mh$ actions from $\alla^\mh$, predicts the next $\mh$ states from $\states^\mh$. We aim to minimize:
\begin{align*}
    \underset{\tau\sim\pi}{E}\left[\hspace{-1.2cm}\sum_{\scriptscriptstyle \hspace{1cm}\substack{s_t,\dots,s_{t+h+1},\\\hspace{.3cm}a_{t*},\dots,a_{t+h*}\in\tau}}\hspace{-1.2cm} \left\vert\left\vert \transition_{\theta}(s_t, a_{t*},\dots,a_{\scriptscriptstyle i+h*}) - [s_t,\dots,s_{\scriptscriptstyle t+h+1}] \right\vert\right\vert^2\right] 
\end{align*}
Where $a_{t*}=\{a_{ti}\}_{i\in\agents}$. In practice, this is achieved by caching trajectories and then training the neural network to minimize the MSE loss. That is, we minimize the loss $\mathcal{L}_\theta^{F}:\mathcal{T}^N\rightarrow\mathbb{R}$:
$$ \frac{1}{\bs\mh}\sum_{\tau \in B}\left|\left| \transition_{\theta}(s_0^\tau, a_{1*}^\tau, \dots, a_{\mh*}) - \left[ s^\tau_1, \dots, s^\tau_\mh \right] \right|\right|^2 $$
Where $B$ represents a batch of trajectories of length $\mh$ sampled uniformly from the cache.

@@

$$\mathcal{L}_\theta^\pi(\batch) + \alpha \frac{1}{\bs\horizon}\sum_{\tau\in\batch}\sum_{a\in\tau} d(a^\tau,(l,r))$$
$$d(x,(l,r)) = \begin{cases}(x-l)^2 & \text{if } x < l \\ (x-r)^2 & \text{if } x > r \\ 0 & \text{otherwise} \end{cases}$$
